---
title: "Data_Organization_2"
output: html_document
date: "2025-05-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### I AM NOT THE ONE WHO CREATED THE FOLLOWING CODE 

### I PROVIDE IT FOR THE BETTER UNDERSTANDING OF THE
### FINAL FILE THAT WAS USED FOR FURTHER ANALYSIS ("DataForAnalysisv1") 

THIS FILE IS USED IN ORDER TO BETTER ASSIGN DATEST TO EACH OF THE OBSERVATIONS,
TO REMOVE DUPLICATES, TO CONVERT NISP COUNTS TO PROPORTIONS AND STANDARDIZE THE 
DATA AND FINALLY, TO EXTRACT REINDEER SUITABILITY

* These dates in our data derived from radiometric, luminescence
and amino acid methods, and electron spin resonance. For contexts with multiple dates, 
we sum the probability estimates and use the year with the highest summed probability.

* To refine interpretive value, proportional abundances were calculated 
by excluding rodent remains, standardizing NISP values across assemblages.
+ removed the proportions summing to greater than 1 and those with NAs as well

* The dataset was paired with modeled estimates of habitat suitability for Rangifer tarandus,
derived from fossil and eDNA records. As these models generate predictions spanning 20–69 ka,
only assemblages within this temporal range could be assigned corresponding suitability values.
Consequently, faunal assemblages dating outside this interval (i.e., from 184 ka to 70 ka) are not
associated with reindeer suitability estimates.


# Determining a Date

Each observation in the Neanderthal data is a context that represents occupation of a site in Eurasia before 20ka BP. To narrow this down to Neanderthals, we will constrain the sample by location (pre-50ka BP) and location (Western Eurasia). This should keep any other Homo species from accidentally making it into our sample.

But first, in order to do this, we need to assign each observation to a date. To do this, we will use the dates table (Supplemental 2-3), which contain results from absolute dating methods, including C14, ESR, OSL, U Series, and TL. Some sites only have one of these, while others have some combination of the above.

In order to estimate date based on a series of observations at any given site context, I sum the probability distribution functions for all dates observed at a site context.

The code below:

- Pulls site i from the Neanderthal data (Supplemental 1),
- Pulls all dates from the three tables that have the same locality, assemblage, and GeoLayer,
- If radiocarbon dates are present, it calibrates the radiocarbon dates,
- If any other dates are present, it estimates a probability density function,
- Next, it sums the probability density functions from the non-c14 dates,
- Then, it creates a summed probability distribution of the radiocarbon dates, if present,
- Then, it sums the non-c14 and c14 probability distribution functions,
- Last, to select a date, it takes the year BP associated with the highest point on the summed probability distribution function.
- Note that this code has already been run.

```{r,eval=F}
maindata<-read.csv("./CleanedData_Zooarch.csv")

#reading in the location data
#These should bind by GeoLayer


#reading in the previous layers
data1<-read.csv("./Dates_Data/Dates_Geog.csv",na.strings="-")
data2<-read.csv("./Dates_Data/Dates_ArchLayer.csv",na.strings="-")
data3<-read.csv("./Dates_Data/Dates_Assemb.csv",na.strings="-")
data3$Age<-as.numeric(data3$Age)
data3$Pos..standard.deviation<-as.numeric(data3$Pos..standard.deviation)

#adding a column for mean date and sd, although its not really a mean, its more the most likely age with the sum of the distribution within 5000 years
maindata$meanage<-NA
maindata$sdage<-NA
#for loop!
for (i in 1:nrow(maindata)){
  #identifying the site
  site<-maindata[i,]
  #pulling observations from the 3 data tables into 1 table
  dates1<-data1[data1$Locality==site$Locality 
        & data1$Assemblage.name==site$Assemblage
        & data1$GeoLayer==site$GeoLayer.Name,]
  dates2<-data2[data2$Locality==site$Locality 
                & data2$Assemblage.name==site$Assemblage
                & data2$GeoLayer==site$GeoLayer.Name,]
  dates3<-data3[data3$Locality==site$Locality 
                & data3$Assemblage.name==site$Assemblage
                & data3$GeoLayer==site$GeoLayer.Name,]
  dates<-rbind(dates1,dates2,dates3)
  #removing observations where the Age is NA
  dates<-dates[!is.na(dates$Age),]
  dates<-dates[!is.na(dates$Pos..standard.deviation),]
  dates<-dates[dates$Pos..standard.deviation<90000,]
  #skipping sites with no observation
  if (nrow(dates)== 0){next}
  
  #Now we will only have dated sites! But we still need to deal with calibrating radiocarbon dates so that they are comparable to other methods.
    c14sites<-dates[dates$Dating.method=="14C (radiocarbon) dating",]
    
    #if there are radiocarbon dates, we calibrate them!
    if (nrow(c14sites)!=0){
    cals<-calibrate(x=c14sites$Age,
                errors=c14sites$Pos..standard.deviation,
                calCurves = rep("intcal20",nrow(c14sites)),
                verbose=F)}
  
  #Now we need to know if there are other dates
  notc14sites<-dates[dates$Dating.method!="14C (radiocarbon) dating",]
  udates<-data.frame(year=20000:200000,pdf=0)
  if (nrow(notc14sites)!=0) {
    #stack the distributions and find the mean and sd
    
    for (z in 1:nrow(notc14sites)){
      udates$pdf<-udates$pdf+dnorm(c(20000:200000),notc14sites$Age[z],notc14sites$Pos..standard.deviation[z])
    }
    meanage<-udates$year[udates$pdf==max(udates$pdf)]
    maindata$meanage[i]<-meanage
    maindata$sdage[i]<-sum(udates$pdf[udates$year<=meanage+5000 | udates$year>=meanage-5000])/nrow(notc14sites)
  } 
  
  if (nrow(c14sites)!=0){
    #stack the c14s only and find the mean and sd
    calspd<-as.data.frame(spd(cals,c(200000,20000))$grid)
    #stack calspd and udates
    calspd$totheigh<-calspd$PrDens+udates$pdf
    
    meanage<-calspd$calBP[calspd$totheigh==max(calspd$totheigh)]
    maindata$meanage[i]<-meanage
    maindata$sdage[i]<-sum(calspd$totheigh[calspd$calBP<=meanage+5000 | calspd$calBP>=meanage-5000])/(nrow(c14sites)+nrow(notc14sites))
    #add to mainDF
  }
  
}

write.csv(maindata,"./KonstantinosDatawDates.csv",row.names = F)
```

# Removing Duplicates

Are they duplicates? Do we want to filter out different contexts from the same site if they have different observations?

```{r}
maindata<-read.csv("./KonstantinosDatawDates.csv",na.strings = c("-","NA"))

maindata$meanage[maindata$meanage==20000 | maindata$meanage==200000]<-NA

#now to remove duplicates!
#these are duplicates in location (long,lat), and time (z). We will need to do this again when they end up binned, I think.
maindata<-distinct(maindata,X,Y,NISP.Sum.class.Mammalia,.keep_all=T)
#and remove NAs
maindata<-maindata[is.na(maindata$meanage)!=T,]

#Trim Temporally - definint the temporal extent
#maindata<-maindata[maindata$meanage<145000,]
#maindata<-maindata[maindata$meanage>50000,]
hist(maindata$meanage,breaks=100) #this almost looks like I could run with 1000 year intervals
```

# Converting NISP Counts to Proportions

```{r,eval=F}
#Removing sites with no NISP totals for mamammals
maindata<-maindata[is.na(maindata$NISP.Sum.class.Mammalia)==FALSE,]
#making rodent NAs 0s
maindata$NISP.Sum.order.Rodentia[is.na(maindata$NISP.Sum.order.Rodentia)==T]<-0

#Subtracting rodents from the mammal nisp
maindata$totalNISP<-maindata$NISP.Sum.class.Mammalia-maindata$NISP.Sum.order.Rodentia

#removing data with 0 NISP
maindata<-maindata[maindata$totalNISP!=0,]

#calculating proportions
for (i in 17:37){
        maindata[,i]<-round(maindata[,i]/maindata$totalNISP,5)
}


#Now to remove out some of the bad data
listofbads<-list()

#clearing out the proportions greater than 1 where NISP of a species was greater than Total NISP...
for (i in 1:nrow(maindata)){
  obs<-maindata[i,]
  #recording the row for those with no values, so just NAs
  if (is.na(sum(obs[,17:37]))==T) {
    append(listofbads,values=i)
  } else if (sum(obs[,17:37])>1) {
    listofbads<-append(listofbads,values=list(i)) #recording the row of those that sum greater than 1...
  }
}

#removing the bads
maindata<-maindata[-as.numeric(listofbads),]
```

# Extracting Reindeer Suitability

These only go back to 69ka ago. So it will only cover a subset of the sites, those between 20ka and 69ka ago.


```{r,eval=F}
setwd(" ... ")    # I DO NOT PROVIDE THIS SPECIFIC DIRECTORY FOR PRIVACY REASONS 
toplist<-list.files(path="./Predictions/MaxEnt/Model3/Continuous/")

RD<-stack(paste0("./Predictions/MaxEnt/Model3/Continuous/",(toplist[c(order(as.numeric(gsub("[^0-9]", "", toplist))))]))[21:length(toplist)])

#rounding observations to centuries
maindata$millie<-round(maindata$meanage,-3)
maindata$RD_Suitability<-NA

mills<-seq(20000,69000,1000)

for (i in 1:length(mills)){
  Thesesites<-maindata[maindata$millie==mills[i],]
  Thismap<-RD[[i]]
  #skipping centuries with no observations
  if (nrow(Thesesites)==0) { next }
  #begin extraction
  Thesesites_sp<-st_as_sf(Thesesites,coords=c("X","Y"),crs=projection(sPDF))
  lambert<-CRS("+init=epsg:3035")
  Thesesites_sp<-st_transform(Thesesites_sp,lambert)
  
  #extract their values from the rasters
  extracted<-as.data.frame(extract(x=Thismap,y=Thesesites_sp))
  
  #add them to our original dataframe
  maindata[maindata$millie==mills[i],]$RD_Suitability<-extracted[,1]
}

write.csv(maindata,"./StudentDataCollection/Konstantinos_Neanderthals/DataForAnalysisv1.csv",row.names = F)

```
